tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2023-11-20 paper.tex
Tweaks

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models~\citep{uffelmann2021}, adapted to the ARG context (Fig~\ref{fig:tree-illustration}).
Each trait is associated with one or more causal sites (position
on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide). Because we are reasoning
about the inheritance of alleles at a site on the genome,
we can consider just the local tree at that site and
the network structure of the ARG is not important~\citep{wong2023general}.
In Fig.~\ref{fig:tree-illustration} we show the local tree
relating three diploid individuals for a given causal site.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.

At a given site, the causal allele is associated with an
effect size $\beta$,
and every node in the tree that carries this allele has a
``genetic value'' of $\beta=0.05$. In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele, and therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
% TODO I find this clearer than saying "additive genetic variance"
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
In the simplest case, simulating a phenotype for an ARG requires
very little code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model, h2=0.3)
\end{lstlisting}
We first create \texttt{model} representing the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function, which also takes the trait model
and narrow sense heritability $h^2$ as parameters.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed by a host of Python data science tools.

>>>>

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
models~\citep{uffelmann2021}, adapted to the ARG context.
Each trait is associated with one or more causal sites
(positions on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide) associated with an effect size $\beta$.
Every node in the local tree that inherits the causal allele
at that site is said to have a ``genetic value'' of $\beta$.
Because we are reasoning
about the inheritance of alleles at a site on the genome,
we can consider just the local tree at that site and
the network structure of the ARG is not important~\citep{wong2023general}.

In Fig.~\ref{fig:tree-illustration} we show an example
tree for three individuals.
Because these individuals are diploids, each is associated
with two nodes in the tree (highlighted by colour).
Ancestral nodes are not associated with individuals here,
but in general the ARG may be embedded in a multigenerational
pedigree, where some internal nodes would be associated
with individuals.
In the example of
Fig.~\ref{fig:tree-illustration}, T is chosen as the causal
allele and $\beta=0.05$, and
therefore all nodes descending from \textsf{i}
have genetic value $0.05$, except \textsf{e} which
has zero because of the back-mutation to A.
Following the standard practise in GWAS~\citep{uffelmann2021},
we assume the additive model such that the overall
genetic value of an individual is the sum of its
nodes' genetic values.
Given these per-individual genetic values, the final phenotype
is then generated by adding some environmental noise.
This noise is simulated from a normal distribution with mean zero
and variance of $V_G(1-h^2)/{h^2}$,
% TODO I find this clearer than saying "additive genetic variance"
where $V_G$ is the variance of the individual genetic values
and $h^2$ is the narrow-sense heritability provided as input by the user.
% The \texttt{tstrait} API is modular, however, and other forms of environmental
% noise can be easily generated by the user, if they wish.

\subsection{Interface}
\texttt{Tstrait} is a Python library, building on the \texttt{tskit}
ARG library~\citep{ralph2020,wong2023general} and the rich
Python data-science ecosystem~\citep{numpy}.
In the simplest case, simulating a phenotype for an ARG requires
very little code:
% NOTE this listing assumes num_causal has been moved, or removed.
% We need to check whether this actually works in the final version.
\begin{lstlisting}[language=Python]
import tstrait as tst
model = tst.trait_model('normal', mean=0, var=1)
result = tst.sim_phenotype(arg, model)
\end{lstlisting}
We first create \texttt{model}, which represents the distribution
from which effect sizes are drawn. Five commonly used
univariate distributions are supported, along with the
multivariate normal distribution to model pleiotropic traits.
Given this model, we can then simulate phenotypes for the individuals
in an ARG (as a  \texttt{tskit TreeSequence}) using the
\texttt{sim\_phenotype} function.
The return value \texttt{result} is an object encapsulating
two Pandas dataframes~\citep{mckinney2010data}: one describing the simulated
effect sizes and the other describing the genetic values,
environmental noise and final phenotypes for each individual.
The simulation results can then be efficiently and conveniently
processed by a host of Python data science tools.

[WIP]
